---
title: "Great Lake Fish Analysis"
author: "Cesia Mendez"
format: html
editor: visual
execute:
  keep-md: true
---

## Great Lake Fish Analysis 

Commercial fish catch data were published by the Great Lakes Fishery Commission in 1962 and covered the period 1867-1960. A supplement covering the years 1961-1968 was released in 1970, and a revised edition covering the years 1867-1977 was published in 1979. This third update of a web-based version covers the period 1867-2015. The intent is to update at approximately five-year intervals. The files are intended for open use by the public. We ask only that the commission be acknowledged when these records are used in presentations and publications.

## Getting Data 

Data will be updated here

```{r}
fishing <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-06-08/fishing.csv')
stocked <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-06-08/stocked.csv')
```

```{r}
head(fishing)
```

```{r}
#install.packages("tidymodels")
library(tidymodels)

my_data_splits <- initial_split(fishing, prop = 0.5)

exploratory_data <- training(my_data_splits)
test_data <- testing(my_data_splits)
```

```{r}
head(test_data)
```

```{r}
tail(test_data)
```

### Hypotheses 

1.  Do the number of species vary by region?
2.  Do the types of species vary by region?
3.  Does the number of species in a region depend on the resources available in that region?
